A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-10-16 main.tex
Lasso -> lasso

One method for analyzing data in the presence of correlated loci is LD pruning, which involves sequentially thinning SNPs, so that only those correlated with each other below some desired threshold are used in modeling and testing procedures. LD pruning can be an important preprocessing step in a GWAS analysis, for which there exist a variety of algorithms and methods. An alternative solution to the problem of LD is the use of multivariate methods such as multiple linear regression, which adjusts for and estimates the effects of numerous variants simultaneously. However, classical approaches rely on collecting data with more observations than unknown parameters. This is typically infeasible in genome-wide studies where the number of features greatly exceeds the number of observations. Penalized regression methods make such problems tractable by constraining the solution space of coefficient estimates. The Lasso is a widely used penalized regression method \citep{tibshirani1996regression}. It relies on an assumption of sparsity, which allows coefficient estimation and variable selection to be accomplished simultaneously. It also provides considerable computational advantages that scale up well to very large data sets. These features make the Lasso attractive for the analysis of high-dimensional genetic data where identifying a list of variants most highly associated with a particular trait is often of interest.

As with multiple linear regression models, Lasso penalized regression assumes independent observations and uncorrelated errors. In the presence of population structure or cryptic relatedness, however, samples are correlated, violating this assumption. If unaccounted for, this dependency can result in the identification of spurious associations. Though methods to correct for population stratification and cryptic relatedness have been extensively investigated, until recently the majority of this work has focused on plant and animal breeding data, much of which is experimentally controlled \citep{Amin2007, hoffman2013correcting, price2006principal, Rakitsch2012, bhatnagar2019simultaneous, Sillanpaeae2011}. While many standard correction techniques may in principal be applied to human genetic data, their performance has not been well-studied within the context of human ancestry and non-experimental conditions \citep{lawson2019population, barton2019population}.

@@

In order to clarify and emphasize the importance of the subtle distinction between population structure and environmental heterogeneity, we present a detailed review of relevant concepts and methods. The remainder of this paper is organized as follows. In Section \ref{sec:background} we provide a summary of the causes and consequences of structure in the genetic data of seemingly unrelated individuals and formally define terminology used throughout this paper. We emphasize that confounding due to population structure, frequently cited as a driver of spurious associations, is a function of environmental heterogeneity, where genetic data serves as a proxy for differential environmental exposures \citep{Sillanpaeae2011, sul2018population, vilhjalmsson2012nature, barton2019population}. In so doing, we refine the definition of confounding due to population structure in terms of a non-genetic mechanism. In addition, we briefly review historical methods of correcting for population stratification and relatedness. In Section \ref{sec:methods} we provide a detailed review of the two most prevalent multivariate methods of adjusting for population stratification and relatedness at present: PCA adjusted Lasso penalized regression (PC-Lasso) and Lasso penalized multivariate LMMs (LMM-Lasso). We review the statistical details of how these methods attempt to correct for population structure and environmental heterogeneity. In Section \ref{sec:results}, we illustrate the concepts described in Section \ref{sec:methods} via simulation studies, where the data-generating mechanism formally differentiates between population structure and environmental heterogeneity. In addition, these results identify scenarios where particular methods may be most and least effective in accurately estimating the effect sizes of observed SNPs in the presence of unobserved confounding.

@@

The remainder of this review will focus on the performance of such multivariate Lasso-penalized methods for the analysis of continuous, normally distributed outcomes. Specifically, PC-adjustment and LMM methods, which we will refer to as PC-Lasso and LMM-Lasso, respectively. We consider two similar implementations of the LMM-Lasso method, one developed by \citet{Rakitsch2012} and a more recent implementation from \citet{bhatnagar2019simultaneous}. Subsequent references to LMM-Lasso apply to both implementations, unless otherwise stated. Where necessary, we will differentiate between these implementations by referring to them as LMM-Lasso-Rakitsch and LMM-Lasso-ggmix, respectively.

@@

To describe and compare PC-Lasso and LMM-Lasso, we consider an $n \times p$ matrix of genotype data $\bG$ for $n$ subjects and $p$ SNPs, where each $G_{ij} \in \{ 0, 1, 2 \}$ enumerates the number of allele copies subject $i$ has for SNP $j$. In order to equitably compare SNP effects, regardless of allele frequency, in practice $\bG$ is standardized. We denote the standardized version of this genotype matrix as $\bX$, where each $X_{ij} = (G_{ij} - 2 \pi_j) / \sqrt{\pi_j (1 - \pi_j)}$ and $\pi_j$ is the minor allele frequency (MAF) of SNP $j$ \citep{zhang2015principal, price2006principal}. Additionally, $\bX_0$ represents an $n \times p_0$ matrix of unpenalized covariates and the intercept. Thus, at a minimum, $\bX_0$ will be an $n \times 1$ vector of 1s to represent the intercept, in which case $\bbeta_0 \equiv \beta_0$ is the scalar intercept. However, $\bX_0$ may additionally include observed covariates such as age, sex, or smoking history, which are included in the model without penalty. We assume that the $n \times 1$ vector of continuous phenotype values, $\by$, can be expressed as the sum of these covariates, additive SNP effects, environmental effects, and observational noise, and consider the linear model,
\begin{equation}
    \label{eqn:model}
    \by = \bX_0 \bbeta_0 + \bX\bbeta + \bZ\bgamma + \beps
\end{equation}
where $\bbeta_0$ is a $p_0 \times 1$ vector which includes the intercept term and the effects of any unpenalized covariates, $\bbeta$ is a $p \times 1$ vector of SNP effects, $\bgamma$ is a $q \times 1$ vector of environmental effects, $\bZ$ is an $n \times q$ matrix that allocates environmental effects among subjects, and $\beps \sim N(\boldsymbol{0}, \sigmaee \bI_n)$. Often it is assumed that subjects belong to $q \le p$ subpopulations, and $\bZ$ is a corresponding incidence matrix of 0s and 1s. However, as we will discuss, this is not the only possibility.

A key feature which differentiates \eqref{eqn:model} from traditional linear mixed models is that in our setting, $\bZ$ is not observed directly, although we still wish to correct for it in order to remove the effects of confounding influences. A common convention is to assume $\bZ = \bX$ \citep{wang2018multiplex, lippert2011fast, yang2014advantages}; we will discuss the rationale for this approach in Section~\ref{sec:sim_env_conf}.  This, in turn, motivates the methods by which LMM-Lasso and PC-Lasso attempt to model unobserved confounding effects using observed genetic data.

Both PC-Lasso and LMM-Lasso rely on matrix decompositions of $\bX$ and the realized relationship matrix (RRM), $\bK = p^{-1} \bX \bX^\T$, which describes the pair-wise genetic similarity among subjects \citep{hayes2009increased}. Using the singular value decomposition (SVD) of the standardized genotype matrix, we can derive the spectral decomposition of the RRM:

@@

%------------------------------%
\subsection{PC-Lasso}
PC-Lasso uses the first $k$ principal components of $\bK$ to approximate $\bZ\bgamma$ and includes them in the model as unpenalized covariates. This corresponds to the regression model
\begin{equation}
    \label{eqn:pca_reg}
    \by = \bX_0\bbeta_0 + \bX\bbeta + \bU_{1:k}\boldsymbol{\alpha} + \beps 
\end{equation}
$$ \beps \sim N(\boldsymbol{0}, \sigmaee \bI), $$
where $\bU_{1:k}$ is the $n \times k$ matrix of principal component vectors, $\boldsymbol{\alpha}$ their corresponding $k \times 1$ coefficient vector, and all other parameters are as defined in equation \eqref{eqn:model}. In a penalized regression framework, this model yields the objective function
\begin{equation}
    \label{eqn:pca_obj}
    Q(\bbeta, \boldsymbol{\alpha}|\bX, \by) = \frac{1}{2n} ||\by - \bX_0\bbeta_0 - \bX \bbeta- \bU_{1:k}\boldsymbol{\alpha}||_2^2 + \lambda || \bbeta ||_1
\end{equation}
where $\lambda$ is a tuning parameter, and $|| \boldsymbol{\beta} ||_1 = \sum_{j=1}^p |\beta_j|$ denotes the $\ell_1$ norm of the regression coefficients, and $||\mathbf{x}||_2 = \sqrt{\sum_{i=1}^p x_i^2}$ the $\ell_2$ norm. Note that the principal component regression coefficients, $\boldsymbol{\alpha}$, are not included in the penalty term.

%------------------------------%
\subsection{LMM-Lasso}
\label{sec:lmmlasso}

Conversely, LMM-Lasso models $\bZ\bgamma$ as a random effect, which corresponds to the LMM model
\begin{equation}
    \label{eqn:lmm_reg}
    \by = \bX_0\bbeta_0 + \bX\bbeta + \bX\bgamma+ \beps,
\end{equation}
$$ \beps \sim N(\boldsymbol{0}, \sigmaee \bI) $$
$$ \bgamma \sim N(\boldsymbol{0}, p^{-1} \sigmagg \bI) \Rightarrow \bX \bgamma \sim N(\boldsymbol{0}, \sigmagg \bK), $$
where $\sigmagg$ is the additive genetic variance, and we assume $\beps \independent \bgamma$. Rather than assuming anything about $\bgamma$ directly, a common representation paramaterizes the model in terms of $\bu=\bX\bgamma$, which yields an equivalent model.

We will first derive the LMM-Lasso objective function in the absence of unpenalized covariates for the sake of notational simplicity. Following equation \eqref{eqn:lmm_reg}, in the absence of unpenalized covariates the LMM assumes $\by \sim N(\bX \bbeta, \bV)$, where $\bV = \sigmagg \textbf{K} + \sigmaee \textbf{I}$. LMM-Lasso uses the covariance structure of the errors and properties of the multivariate normal distribution to transform the original data and diagonalize the errors in a manner akin to generlized least squares (GLS). Specifically, left-multiplying the original data by $\bV^{-1/2}$ yields

\begin{equation}
\label{eqn:rtY_distn}
\bV^{-\frac{1}{2}}\by \sim N(\bV^{-\frac{1}{2}} \bX \bbeta, \bI),
\end{equation}
and thus an ordinary Lasso model can be fit on the transformed data, ($\bV^{-\frac{1}{2}}\bX, \bV^{-\frac{1}{2}}\by$).

@@

Different variance component estimation methods lead to different forms of $\widehat{\bW}$ and two distinct LMM-Lasso variants: LMM-Lasso-Rakitsch and LMM-Lasso-ggmix. LMM-Lasso-Rakitsch uses a two-step procedure for estimating the variance components, $\sigmaee$ and $\delta := \sigmaee / \sigmagg$ \citep{Rakitsch2012}. These variance components and the corresponding weights are estimated once under the assumption of a null model. A Lasso model can then be fit on the transformed data to estimate $\boldsymbol{\beta}$ using standard software for coordinate descent, such as \texttt{glmnet} \citep{glmnet} or \texttt{ncvreg} \citep{ncvreg}. LMM-Lasso-ggmix iteratively updates $\widehat{\boldsymbol{\beta}}$ and the variance components via a block coordinate descent algorithm \citep{bhatnagar2019simultaneous}. This iterative update scheme prevents the use of standard Lasso software, but is implemented in the \texttt{R} package \texttt{ggmix} \citep{ggmix}. We further discuss differences between these methods in Section \ref{sec:results}.

@@

%------------------------------%
\subsection{Relationship between PC-Lasso and LMM-Lasso}

It has been shown that the random effect formulation of model \eqref{eqn:model} used by LMM-Lasso can be equivalently expressed in terms of fixed effects, in the absence of penalization. \citet{zhang2015principal} derived this equivalence using a probabilistic PCA formulation, while \citet{hoffman2013correcting} used the singular value decomposition of $\bX$, which we briefly review.

@@

Whether modeled as a fixed or random effect, the underlying effect of $\bZ \bgamma$ on phenotype remains the same, but its treatment as a fixed or random effect has important implications in terms of model fitting and estimation. Modeling unobserved confounding using fixed effects requires estimation of $p_0 + p + k$ parameters, compared to the $p_0 + p + 2$ required of an LMM, which may result in a loss of power \citep{zhang2015principal}. Additionally, when modeling $\bZ \bgamma$ using fixed effects, determining the appropriate number of principal components, $k$, to adjust for is non-trivial, and has been the subject of numerous studies \citep{patterson2006population, zhao2018practical}. It has been shown that including too many or too few principal components can result in power loss or increased Type I errors, respectively \citep{zhang2015principal}. Nevertheless, a common practice is to include the ten largest principal components in the analysis \citep{zhao2018practical}. This is the approach we take when implementing PC-Lasso in Section \ref{sec:results}.

@@

For each $\bX$ matrix, $\bZ$ was a corresponding incidence matrix of 1s and 0s identifying subpopulation, or founding population membership, which assigns the corresponding environmental effect found in $\bgamma$. Various $\bgamma$ patterns were considered in order to investigate the effects of distinct forms of environmental confounding. Our initial hypothesis was that simple patterns of environmental effects could be well-approximated by a linear combination of the observed genetic data and well-corrected for by both PC-Lasso and LMM-Lasso, while PC-Lasso would be unable to represent and correct for more complex patterns of environmental effects.

@@

Although cross-validation and information criterion-based methods are commonly used to select $\lambda$ values, in this context they have the potential to result in misleading comparisons across methods. Cross-validation may be inadvisable for LMMs, and in the presence of samples with underlying structure, as is the case here \citep{roberts2017cross}. While information criterion methods such as BIC present another option for selecting $\lambda$, they may be unstable in high dimensions. In addition, because Lasso, PC-Lasso, and LMM-Lasso use different likelihoods, an information-criterion-based approach would not lend itself to comparable results across methods.

@@

%------------------------------%
\subsection{PC-Lasso and LMM-Lasso outperform the ordinary Lasso in the presence of environmental confounding}
%------------------------------%

We begin by comparing the ordinary Lasso, PC-Lasso, and LMM-Lasso as the level of environmental confounding is varied. Estimation accuracy for Lasso, PC-Lasso, and LMM-Lasso was evaluated based on mean squared error (MSE) \anna{MSPE sims running - maybe add here}. Figure \ref{fig:mse} presents these results for all three data types in a 1:1 SNR setting, where $\eta = 0.5$. The top row of Figure \ref{fig:mse} shows that when $\xi = 0$ (i.e., no environmental confounding), Lasso and LMM-Lasso perform comparably. PC-Lasso results in a slight inflation in MSE compared to the other methods in the simulated data sets due to the fact that 10 additional predictors (PCs) have been added to the model, but none of them actually explain any variability in the outcome.

@@

The bottom row of Figure \ref{fig:mse} shows the MSE when $\xi = 0.8$. In this case, both PC-Lasso and LMM-Lasso clearly outperform the ordinary Lasso. This improvement is more substantial in the Empirical and Independent Subpopulations data than in the 1D Linear Admixture data. This is due to the fact that the Independent Subpopulations and Empirical data sets are characterized by nonoverlapping genetic subpopulations, meaning that a greater proportion of the environmental effect is in the column space of $\boldsymbol{P}_{\bX}$. We can quantify the proportion of the environmental component in the column space of $\boldsymbol{P}_{\bX}$ for each data type using the CP metric defined in \eqref{eqn:ratio} of Section \ref{sec:sim_env_conf}. Here, we have CP $= 0.47, 0.99$ for the 1D Linear Admixture and Independent Subpopulations, respectively. The large CP of the Independent Subpopulations model reflects the extreme confounding and estimation bias for this data, for which the PC-Lasso and LMM-Lasso methods are able to correct. Meanwhile, the less dramatic improvement in MSE for the 1D Linear Admixture data by the PC-Lasso and LMM-Lasso methods confirms what we know from this data's substantially smaller CP value; that there is less confounding and bias which can be corrected for.

@@

In addition to MSE, these methods' performance was also assessed in terms of variable selection using the true positive rate (TPR) for the first 50 variables each method allowed into the model. Table \ref{tab:var_sel} displays the average TPR for the same settings as in Figure \ref{fig:mse}. The TPRs echo the MSE results in that Lasso and LMM-Lasso perform comparably in the absence of environmental confounding across all data types, while the addition of the 10 unpenalized PCs leads to a slight decrease in TPR in this setting for the 1D Linear Admixture and Independent Subpopulations data sets. In the presence of environmental confounding, both LMM-Lasso and PC-Lasso achieve substantial improvements in TPR over the ordinary Lasso. As with the MSE data, these improvements are more substantial in the Independent Subpopulations and Empirical data sets than in the 1D Linear Admixture data.

\begin{table}[H]
\centering
\begin{tabularx}{400pt}{cclYYY}
\toprule
$\eta$ & $\xi$ & \multicolumn{1}{c}{Method} & 1D Linear Admixture & Independent Subpopulations & Empirical \\ 
\midrule
\multirow{6}{*}{0.5} & \multirow{3}{*}{0} & Lasso & 0.25 (0.05) & 0.23 (0.05) & 0.37 (0.06) \\ 
& & PC-Lasso & 0.23 (0.05) & 0.21 (0.05) & 0.36 (0.06) \\ 
& & LMM-Lasso & 0.25 (0.05) & 0.23 (0.05) & 0.41 (0.06) \\ 
\cmidrule{2-6}
& \multirow{3}{*}{0.8} & Lasso & 0.24 (0.05) & 0.26 (0.06) & 0.29 (0.06) \\
& & PC-Lasso & 0.28 (0.05) & 0.36 (0.06) & 0.46 (0.06) \\ 
& & LMM-Lasso & 0.28 (0.05) & 0.39 (0.06) & 0.57 (0.05) \\
\bottomrule
\end{tabularx}
\caption{True positive rates, mean (SD), for different data types in the presence and absence of environmental confounding with coarse subpopulation structure and dichotomous-discordant environmental effects.}
\label{tab:var_sel}
\end{table}

%------------------------------%
\subsection{LMM-Lasso outperforms PC-Lasso for finer population structures}
%------------------------------%

Given the superior performance of PC-Lasso and LMM-Lasso compared to the ordinary Lasso in the presence of environmental confounding, we next compare these methods in the presence of fine and coarse subpopulation structures as defined in Section \ref{sec:subpopulation_structure}. Synthetic phenotypes were simulated using a 1:1 SNR, $\eta = 0.5$, and varying levels of environmental confounding, $\xi \in \{0.2, 0.5,0.8\}$.

>>>>

One method for analyzing data in the presence of correlated loci is LD pruning, which involves sequentially thinning SNPs, so that only those correlated with each other below some desired threshold are used in modeling and testing procedures. LD pruning can be an important preprocessing step in a GWAS analysis, for which there exist a variety of algorithms and methods. An alternative solution to the problem of LD is the use of multivariate methods such as multiple linear regression, which adjusts for and estimates the effects of numerous variants simultaneously. However, classical approaches rely on collecting data with more observations than unknown parameters. This is typically infeasible in genome-wide studies where the number of features greatly exceeds the number of observations. Penalized regression methods make such problems tractable by constraining the solution space of coefficient estimates. The lasso is a widely used penalized regression method \citep{tibshirani1996regression}. It relies on an assumption of sparsity, which allows coefficient estimation and variable selection to be accomplished simultaneously. It also provides considerable computational advantages that scale up well to very large data sets. These features make the lasso attractive for the analysis of high-dimensional genetic data where identifying a list of variants most highly associated with a particular trait is often of interest.

As with multiple linear regression models, lasso penalized regression assumes independent observations and uncorrelated errors. In the presence of population structure or cryptic relatedness, however, samples are correlated, violating this assumption. If unaccounted for, this dependency can result in the identification of spurious associations. Though methods to correct for population stratification and cryptic relatedness have been extensively investigated, until recently the majority of this work has focused on plant and animal breeding data, much of which is experimentally controlled \citep{Amin2007, hoffman2013correcting, price2006principal, Rakitsch2012, bhatnagar2019simultaneous, Sillanpaeae2011}. While many standard correction techniques may in principal be applied to human genetic data, their performance has not been well-studied within the context of human ancestry and non-experimental conditions \citep{lawson2019population, barton2019population}.

@@

In order to clarify and emphasize the importance of the subtle distinction between population structure and environmental heterogeneity, we present a detailed review of relevant concepts and methods. The remainder of this paper is organized as follows. In Section \ref{sec:background} we provide a summary of the causes and consequences of structure in the genetic data of seemingly unrelated individuals and formally define terminology used throughout this paper. We emphasize that confounding due to population structure, frequently cited as a driver of spurious associations, is a function of environmental heterogeneity, where genetic data serves as a proxy for differential environmental exposures \citep{Sillanpaeae2011, sul2018population, vilhjalmsson2012nature, barton2019population}. In so doing, we refine the definition of confounding due to population structure in terms of a non-genetic mechanism. In addition, we briefly review historical methods of correcting for population stratification and relatedness. In Section \ref{sec:methods} we provide a detailed review of the two most prevalent multivariate methods of adjusting for population stratification and relatedness at present: PCA adjusted lasso penalized regression (PC-lasso) and lasso penalized multivariate LMMs (LMM-lasso). We review the statistical details of how these methods attempt to correct for population structure and environmental heterogeneity. In Section \ref{sec:results}, we illustrate the concepts described in Section \ref{sec:methods} via simulation studies, where the data-generating mechanism formally differentiates between population structure and environmental heterogeneity. In addition, these results identify scenarios where particular methods may be most and least effective in accurately estimating the effect sizes of observed SNPs in the presence of unobserved confounding.

@@

The remainder of this review will focus on the performance of such multivariate lasso-penalized methods for the analysis of continuous, normally distributed outcomes. Specifically, PC-adjustment and LMM methods, which we will refer to as PC-lasso and LMM-lasso, respectively. We consider two similar implementations of the LMM-lasso method, one developed by \citet{Rakitsch2012} and a more recent implementation from \citet{bhatnagar2019simultaneous}. Subsequent references to LMM-lasso apply to both implementations, unless otherwise stated. Where necessary, we will differentiate between these implementations by referring to them as LMM-lasso-Rakitsch and LMM-lasso-ggmix, respectively.

@@

To describe and compare PC-lasso and LMM-lasso, we consider an $n \times p$ matrix of genotype data $\bG$ for $n$ subjects and $p$ SNPs, where each $G_{ij} \in \{ 0, 1, 2 \}$ enumerates the number of allele copies subject $i$ has for SNP $j$. In order to equitably compare SNP effects, regardless of allele frequency, in practice $\bG$ is standardized. We denote the standardized version of this genotype matrix as $\bX$, where each $X_{ij} = (G_{ij} - 2 \pi_j) / \sqrt{\pi_j (1 - \pi_j)}$ and $\pi_j$ is the minor allele frequency (MAF) of SNP $j$ \citep{zhang2015principal, price2006principal}. Additionally, $\bX_0$ represents an $n \times p_0$ matrix of unpenalized covariates and the intercept. Thus, at a minimum, $\bX_0$ will be an $n \times 1$ vector of 1s to represent the intercept, in which case $\bbeta_0 \equiv \beta_0$ is the scalar intercept. However, $\bX_0$ may additionally include observed covariates such as age, sex, or smoking history, which are included in the model without penalty. We assume that the $n \times 1$ vector of continuous phenotype values, $\by$, can be expressed as the sum of these covariates, additive SNP effects, environmental effects, and observational noise, and consider the linear model,
\begin{equation}
    \label{eqn:model}
    \by = \bX_0 \bbeta_0 + \bX\bbeta + \bZ\bgamma + \beps
\end{equation}
where $\bbeta_0$ is a $p_0 \times 1$ vector which includes the intercept term and the effects of any unpenalized covariates, $\bbeta$ is a $p \times 1$ vector of SNP effects, $\bgamma$ is a $q \times 1$ vector of environmental effects, $\bZ$ is an $n \times q$ matrix that allocates environmental effects among subjects, and $\beps \sim N(\boldsymbol{0}, \sigmaee \bI_n)$. Often it is assumed that subjects belong to $q \le p$ subpopulations, and $\bZ$ is a corresponding incidence matrix of 0s and 1s. However, as we will discuss, this is not the only possibility.

A key feature which differentiates \eqref{eqn:model} from traditional linear mixed models is that in our setting, $\bZ$ is not observed directly, although we still wish to correct for it in order to remove the effects of confounding influences. A common convention is to assume $\bZ = \bX$ \citep{wang2018multiplex, lippert2011fast, yang2014advantages}; we will discuss the rationale for this approach in Section~\ref{sec:sim_env_conf}.  This, in turn, motivates the methods by which LMM-lasso and PC-lasso attempt to model unobserved confounding effects using observed genetic data.

Both PC-lasso and LMM-lasso rely on matrix decompositions of $\bX$ and the realized relationship matrix (RRM), $\bK = p^{-1} \bX \bX^\T$, which describes the pair-wise genetic similarity among subjects \citep{hayes2009increased}. Using the singular value decomposition (SVD) of the standardized genotype matrix, we can derive the spectral decomposition of the RRM:

@@

%------------------------------%
\subsection{PC-lasso}
PC-lasso uses the first $k$ principal components of $\bK$ to approximate $\bZ\bgamma$ and includes them in the model as unpenalized covariates. This corresponds to the regression model
\begin{equation}
    \label{eqn:pca_reg}
    \by = \bX_0\bbeta_0 + \bX\bbeta + \bU_{1:k}\boldsymbol{\alpha} + \beps 
\end{equation}
$$ \beps \sim N(\boldsymbol{0}, \sigmaee \bI), $$
where $\bU_{1:k}$ is the $n \times k$ matrix of principal component vectors, $\boldsymbol{\alpha}$ their corresponding $k \times 1$ coefficient vector, and all other parameters are as defined in equation \eqref{eqn:model}. In a penalized regression framework, this model yields the objective function
\begin{equation}
    \label{eqn:pca_obj}
    Q(\bbeta, \boldsymbol{\alpha}|\bX, \by) = \frac{1}{2n} ||\by - \bX_0\bbeta_0 - \bX \bbeta- \bU_{1:k}\boldsymbol{\alpha}||_2^2 + \lambda || \bbeta ||_1
\end{equation}
where $\lambda$ is a tuning parameter, and $|| \boldsymbol{\beta} ||_1 = \sum_{j=1}^p |\beta_j|$ denotes the $\ell_1$ norm of the regression coefficients, and $||\mathbf{x}||_2 = \sqrt{\sum_{i=1}^p x_i^2}$ the $\ell_2$ norm. Note that the principal component regression coefficients, $\boldsymbol{\alpha}$, are not included in the penalty term.

%------------------------------%
\subsection{LMM-lasso}
\label{sec:lmmlasso}

Conversely, LMM-lasso models $\bZ\bgamma$ as a random effect, which corresponds to the LMM model
\begin{equation}
    \label{eqn:lmm_reg}
    \by = \bX_0\bbeta_0 + \bX\bbeta + \bX\bgamma+ \beps,
\end{equation}
$$ \beps \sim N(\boldsymbol{0}, \sigmaee \bI) $$
$$ \bgamma \sim N(\boldsymbol{0}, p^{-1} \sigmagg \bI) \Rightarrow \bX \bgamma \sim N(\boldsymbol{0}, \sigmagg \bK), $$
where $\sigmagg$ is the additive genetic variance, and we assume $\beps \independent \bgamma$. Rather than assuming anything about $\bgamma$ directly, a common representation paramaterizes the model in terms of $\bu=\bX\bgamma$, which yields an equivalent model.

We will first derive the LMM-lasso objective function in the absence of unpenalized covariates for the sake of notational simplicity. Following equation \eqref{eqn:lmm_reg}, in the absence of unpenalized covariates the LMM assumes $\by \sim N(\bX \bbeta, \bV)$, where $\bV = \sigmagg \textbf{K} + \sigmaee \textbf{I}$. LMM-lasso uses the covariance structure of the errors and properties of the multivariate normal distribution to transform the original data and diagonalize the errors in a manner akin to generlized least squares (GLS). Specifically, left-multiplying the original data by $\bV^{-1/2}$ yields

\begin{equation}
\label{eqn:rtY_distn}
\bV^{-\frac{1}{2}}\by \sim N(\bV^{-\frac{1}{2}} \bX \bbeta, \bI),
\end{equation}
and thus an ordinary lasso model can be fit on the transformed data, ($\bV^{-\frac{1}{2}}\bX, \bV^{-\frac{1}{2}}\by$).

@@

Different variance component estimation methods lead to different forms of $\widehat{\bW}$ and two distinct LMM-lasso variants: LMM-lasso-Rakitsch and LMM-lasso-ggmix. LMM-lasso-Rakitsch uses a two-step procedure for estimating the variance components, $\sigmaee$ and $\delta := \sigmaee / \sigmagg$ \citep{Rakitsch2012}. These variance components and the corresponding weights are estimated once under the assumption of a null model. A lasso model can then be fit on the transformed data to estimate $\boldsymbol{\beta}$ using standard software for coordinate descent, such as \texttt{glmnet} \citep{glmnet} or \texttt{ncvreg} \citep{ncvreg}. LMM-lasso-ggmix iteratively updates $\widehat{\boldsymbol{\beta}}$ and the variance components via a block coordinate descent algorithm \citep{bhatnagar2019simultaneous}. This iterative update scheme prevents the use of standard lasso software, but is implemented in the \texttt{R} package \texttt{ggmix} \citep{ggmix}. We further discuss differences between these methods in Section \ref{sec:results}.

@@

%------------------------------%
\subsection{Relationship between PC-lasso and LMM-lasso}

It has been shown that the random effect formulation of model \eqref{eqn:model} used by LMM-lasso can be equivalently expressed in terms of fixed effects, in the absence of penalization. \citet{zhang2015principal} derived this equivalence using a probabilistic PCA formulation, while \citet{hoffman2013correcting} used the singular value decomposition of $\bX$, which we briefly review.

@@

Whether modeled as a fixed or random effect, the underlying effect of $\bZ \bgamma$ on phenotype remains the same, but its treatment as a fixed or random effect has important implications in terms of model fitting and estimation. Modeling unobserved confounding using fixed effects requires estimation of $p_0 + p + k$ parameters, compared to the $p_0 + p + 2$ required of an LMM, which may result in a loss of power \citep{zhang2015principal}. Additionally, when modeling $\bZ \bgamma$ using fixed effects, determining the appropriate number of principal components, $k$, to adjust for is non-trivial, and has been the subject of numerous studies \citep{patterson2006population, zhao2018practical}. It has been shown that including too many or too few principal components can result in power loss or increased Type I errors, respectively \citep{zhang2015principal}. Nevertheless, a common practice is to include the ten largest principal components in the analysis \citep{zhao2018practical}. This is the approach we take when implementing PC-lasso in Section \ref{sec:results}.

@@

For each $\bX$ matrix, $\bZ$ was a corresponding incidence matrix of 1s and 0s identifying subpopulation, or founding population membership, which assigns the corresponding environmental effect found in $\bgamma$. Various $\bgamma$ patterns were considered in order to investigate the effects of distinct forms of environmental confounding. Our initial hypothesis was that simple patterns of environmental effects could be well-approximated by a linear combination of the observed genetic data and well-corrected for by both PC-lasso and LMM-lasso, while PC-lasso would be unable to represent and correct for more complex patterns of environmental effects.

@@

Although cross-validation and information criterion-based methods are commonly used to select $\lambda$ values, in this context they have the potential to result in misleading comparisons across methods. Cross-validation may be inadvisable for LMMs, and in the presence of samples with underlying structure, as is the case here \citep{roberts2017cross}. While information criterion methods such as BIC present another option for selecting $\lambda$, they may be unstable in high dimensions. In addition, because lasso, PC-lasso, and LMM-lasso use different likelihoods, an information-criterion-based approach would not lend itself to comparable results across methods.

@@

%------------------------------%
\subsection{PC-lasso and LMM-lasso outperform the ordinary lasso in the presence of environmental confounding}
%------------------------------%

We begin by comparing the ordinary lasso, PC-lasso, and LMM-lasso as the level of environmental confounding is varied. Estimation accuracy for lasso, PC-lasso, and LMM-lasso was evaluated based on mean squared error (MSE) \anna{MSPE sims running - maybe add here}. Figure \ref{fig:mse} presents these results for all three data types in a 1:1 SNR setting, where $\eta = 0.5$. The top row of Figure \ref{fig:mse} shows that when $\xi = 0$ (i.e., no environmental confounding), lasso and LMM-lasso perform comparably. PC-lasso results in a slight inflation in MSE compared to the other methods in the simulated data sets due to the fact that 10 additional predictors (PCs) have been added to the model, but none of them actually explain any variability in the outcome.

@@

The bottom row of Figure \ref{fig:mse} shows the MSE when $\xi = 0.8$. In this case, both PC-lasso and LMM-lasso clearly outperform the ordinary lasso. This improvement is more substantial in the Empirical and Independent Subpopulations data than in the 1D Linear Admixture data. This is due to the fact that the Independent Subpopulations and Empirical data sets are characterized by nonoverlapping genetic subpopulations, meaning that a greater proportion of the environmental effect is in the column space of $\boldsymbol{P}_{\bX}$. We can quantify the proportion of the environmental component in the column space of $\boldsymbol{P}_{\bX}$ for each data type using the CP metric defined in \eqref{eqn:ratio} of Section \ref{sec:sim_env_conf}. Here, we have CP $= 0.47, 0.99$ for the 1D Linear Admixture and Independent Subpopulations, respectively. The large CP of the Independent Subpopulations model reflects the extreme confounding and estimation bias for this data, for which the PC-lasso and LMM-lasso methods are able to correct. Meanwhile, the less dramatic improvement in MSE for the 1D Linear Admixture data by the PC-lasso and LMM-lasso methods confirms what we know from this data's substantially smaller CP value; that there is less confounding and bias which can be corrected for.

@@

In addition to MSE, these methods' performance was also assessed in terms of variable selection using the true positive rate (TPR) for the first 50 variables each method allowed into the model. Table \ref{tab:var_sel} displays the average TPR for the same settings as in Figure \ref{fig:mse}. The TPRs echo the MSE results in that lasso and LMM-lasso perform comparably in the absence of environmental confounding across all data types, while the addition of the 10 unpenalized PCs leads to a slight decrease in TPR in this setting for the 1D Linear Admixture and Independent Subpopulations data sets. In the presence of environmental confounding, both LMM-lasso and PC-lasso achieve substantial improvements in TPR over the ordinary lasso. As with the MSE data, these improvements are more substantial in the Independent Subpopulations and Empirical data sets than in the 1D Linear Admixture data.

\begin{table}[H]
\centering
\begin{tabularx}{400pt}{cclYYY}
\toprule
$\eta$ & $\xi$ & \multicolumn{1}{c}{Method} & 1D Linear Admixture & Independent Subpopulations & Empirical \\ 
\midrule
\multirow{6}{*}{0.5} & \multirow{3}{*}{0} & lasso & 0.25 (0.05) & 0.23 (0.05) & 0.37 (0.06) \\ 
& & PC-lasso & 0.23 (0.05) & 0.21 (0.05) & 0.36 (0.06) \\ 
& & LMM-lasso & 0.25 (0.05) & 0.23 (0.05) & 0.41 (0.06) \\ 
\cmidrule{2-6}
& \multirow{3}{*}{0.8} & lasso & 0.24 (0.05) & 0.26 (0.06) & 0.29 (0.06) \\
& & PC-lasso & 0.28 (0.05) & 0.36 (0.06) & 0.46 (0.06) \\ 
& & LMM-lasso & 0.28 (0.05) & 0.39 (0.06) & 0.57 (0.05) \\
\bottomrule
\end{tabularx}
\caption{True positive rates, mean (SD), for different data types in the presence and absence of environmental confounding with coarse subpopulation structure and dichotomous-discordant environmental effects.}
\label{tab:var_sel}
\end{table}

%------------------------------%
\subsection{LMM-lasso outperforms PC-lasso for finer population structures}
%------------------------------%

Given the superior performance of PC-lasso and LMM-lasso compared to the ordinary lasso in the presence of environmental confounding, we next compare these methods in the presence of fine and coarse subpopulation structures as defined in Section \ref{sec:subpopulation_structure}. Synthetic phenotypes were simulated using a 1:1 SNR, $\eta = 0.5$, and varying levels of environmental confounding, $\xi \in \{0.2, 0.5,0.8\}$.

\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.9]{figures/mse_diff_subpops.png}
    \caption{Relative performance of LMM-lasso, PC-lasso for fine and coarse subpopulation structures and varying levels of environmental confounding with Independent Subpopulations data, $\eta = 0.5$, and dichotomous-discordant environmental effects. }
    \label{fig:big_vs_small}
\end{figure}
